Higher-Order unification for free!

### 2024-04-11 paper/main.tex
update section map bijection

\begin{theorem}[Fidelity in \maybeeta]
  The introduction 
  of \linketa guarantees \fullref{prop:fidelity} 
  \label{lemma:fidelity-maybeeta}
\end{theorem}

@@

%\section{Enforcing inv.: \NoCaseChange{\nameref{inv:uvaarity}}}
\section{Making \mapStore a bijection}
\label{sec:invariant1}
\todo{l'esempio è sparito}
We report here the problem given in \cref{sec:basic-comp-limitations} where
$X$ is used with two different arities and the output of the compilation does
not respect \cref{inv:map-bijective}. Moreover, merging the two mappings
for $s$ would break \fullref{inv:uvaarity}.
In this section we explain how to replace the duplicate mapping with
\linketa in order to restore the invariants.
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}

We see that the \elpiIn{maybe-eta} as identified $\lambda xy.X \appsep y \appsep
x$ and $\lambda x.f\appsep (X \appsep x) \appsep x$ and the compiler has
replaced them with $A$ and $D$ respectively.
% $A$ is linked with $\lambda x.B_x$, $B$
% has arity $1$ and is $\eta$-linked with $\lambda y.C\appsep y \appsep x$ and $D$
% is linked to the term $\lambda x.f \appsep E_{x}~x$. 
However, the mapping \mapStore breaks \cref{inv:map-bijective}: the \Fo{}
variable $X$ is mapped to two different \Ho variables. To address this problem
we adjust the compiler's output with a \elpiIn{map-deduplication} procedure.

@@

\newcommand{\mapdeduplication}{\emph{map-deduplication}}

@@

@@

% After unification of the two terms, $X$ is assigned to $\lambda x.\lambda y.x$.
% This assignment makes $l_2$ to progress since the \lhs is materialized and by
% unification, between $X$ and $\lambda x.Y_{x}$, $Y_x$ is instantiate to $\lambda
% y.x$. Once $Y_x$ is instantiated, $l_1$ can progress, and set $H_{xy}$ to $x$.
% After all these progresses, $l_1$ and $l_2$ are remove from \linkStore and the
% \elpiIn{progress} fixpoint terminates. Next, the second unification problem is
% run, and $Z$ is set to $f\appsep(\lambda x.x)$. This unification wakes up $l_3$
% and since $Z$ starts with the \elpiIn{app} node, the $\eta$-expanded version of
% $Z$ is unified with $\lambda x.f \appsep G_{x}~x$ and $G_x$ is set to $x$.
% As last step, the last link is progressed and the final \Ho substitution is
% $\{X _{} \mapsto \lambda x.\lambda y.x, Y _{x} \mapsto \lambda y.x, 
% G _{yx} \mapsto y, Z _{} \mapsto f~\lambda x.x, 
% H _{x} \mapsto \lambda y.y\}$.
% The decompilation phase is only charged, in this example to solve the mappings,
% since no suspended links remain. The only mapping in the list is
% \mapping{F}{H}{2}, which will assign the $F$ variable in \Fo{} to $\lambda xy.y$

% \noteE[inline]{dire che preserviamo l'invariante che tutte le variable sono fully-applied
% }
\section{Handling of \notllambda}\label{sec:beta}

>>>>

\begin{theorem}[Fidelity in \maybeeta]
  Given a list of unification problems \foUnifPb, such that 
  $\forall t, t\in\subterm{\foUnifPb} \land t \notin \notllambda$,
  the introduction 
  of \linketa guarantees \fullref{prop:fidelity}.
  \footnote{We also suppose that any higher-order variable is always applied with
  the same number of arguments. This problem is addressed in \cref{sec:invariant1}}
  \label{lemma:fidelity-maybeeta}
\end{theorem}

@@

%\section{Enforcing inv.: \NoCaseChange{\nameref{inv:uvaarity}}}
\section{Making \mapStore a bijection}
\label{sec:invariant1}

In \cref{sec:grounwork}, we introduced the definition of ``memory map'' (\mapStore).
This memory allows to decompile the \Ho terms back to the object language.
It is the case that, while solving unification problems, a same unification
variable $X$ is used multiple times with different arities.
%
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}

In the unification problems \foUnifPb above, we see that $X$ is used with arity $2$ in
$\foUnifPb_1$ and with arity $1$ in $\foUnifPb_2$. 
By \fullref{inv:uvaarity}, we are not allowed to use a same \Ho variable to 
represent the two occurrences of $X$.
If we execute \hrun, we remark that the unification fails.
There is in fact a major problem:
\hstep is not conscious of the connection between the variables $C$ and $E$ (both corresponding to $X$),
since no link in \linkStore puts $C$ and $E$ in relation and decompilation
does not work properly if a \Fo{} variable is mapped to two distinct \Ho variables. 
The two main drawbacks connected to this situation are 
firstly the lost of \fullref{prop:fidelity} and
secondly, if we want to guarantee at least \fullref{prop:simulation}, we should
overcomplicate the decompilation phase. In order to ease the second
drawback, we pose the following property:

\begin{proposition}[\mapStore is a bijection]
  Given a list of unification problems \foUnifPb, then the memory map \mapStore
  compiled from \foUnifPb is a bijection relating the \Fo{} and the \Ho{} 
  variables.
\end{proposition}

We finally adjust the compiler's output with a \elpiIn{map-deduplication} procedure.

@@

\newcommand{\mapdeduplication}{\emph{map-deduplication}\xspace}

@@

\begin{theorem}[Fidelity with \mapdeduplication]
  Given a list of unification problems \foUnifPb, such that 
  $\forall t, t\in\subterm{\foUnifPb} \Rightarrow \wellb(t) \lor t \in \maybeeta$,
  if \foUnifPb contains two same \Fo{} variables with different arities,
  then \mapdeduplication guarantees \fullref{prop:fidelity}
\end{theorem}

\begin{proof}[Proof sketch]
  By the definition of \mapdeduplication, any two same \Fo{} variables $X_1, X_2$
  with different arities are related with \linketa. 
  If one of the two variables is instantiated, the corresponding
  \linketa is triggered instantiating the related variable.
  This allows to make unification fail if $X_1$ and $X_2$ are unified with 
  different terms. Finally, since \foUnifPb contains only terms thar are
  either \wellb or \maybeeta, by \cref{lemma:fidelity-maybeeta},
  we can conclude the proof.
\end{proof}

@@

\section{Handling of \notllambda}\label{sec:beta}
